Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2020-12-18 manuscript.tex
Updates from Overleaf

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA}. Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}. Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}. Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.

@@

In this work, we study the same four sequences explored by Sanstead et al. in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}. We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}. We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent. We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.

% this section should look more like a conclusion foreshadowing conclusions and main take aways 
We find that 3spn2 allows us to access long length and time scales and observe sequence-dependent kinetics and relaxation rates that are, within a corrective scaling factor, in agreement with observed experimental trends. Applying SRV-MSMs enables the construction of high-resolution sequence-dependent kinetic models of hybridization and dissociation. Consistent with previous studies \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity}, we find the degree of repetitiveness in the sequence and therefore the kinetic accessibility and thermodynamic stability of shifted states leads to richer folding dynamics populated by a diversity of long-lived metastable states. Disrupting repetitive stretches of bases by judicious placement of interrupting G:C pairs can tune the landscape from rich 6-state to very simple 2-state (“all-or-nothing”) behavior. Furthermore, concentrating G:C pairs at the center of a duplex produces a longer-live frayed state that facilitates the hybridization and dissociation process. Analysis within metastable state populations using diffusion maps allows us to explore the conformational diversity and heterogeneity within each state at time scales shorter than the kinetic resolution of the SRV-MSMs. This provides us with molecular-level understanding for the differential stability of various conformational states as a function of oligomer sequence. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.

@@

We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm \citep{Sanstead2016, Phys2014}. We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff. We ran our simulations in the NVT ensemble. Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent. \citep{Schneider1978Molecular-dynamicsTransitions}. Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states. We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequence, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation. Half of runs were initialized from the native hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. We observed between 55-100 hybridization/dissociation events for each of the four sequences.

@@

@@

% change state names and scientific notation
% include D here? can see what it would look like...
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=160mm, 
        scale=1]{Fig3.pdf}
        \caption{Nearest neighbor representations, molecular renderings, and sequence-dependent probabilities for PCCA+ macrostates. Shifted state abbreviations indicate the direction of the shifted overhang (5' vs. 3') and the number of shifted motifs relative the native state (1 vs. 2).}
        \label{fig:allseq_table}
	\end{center}
\end{figure}

% potentially get rid of State representation, ask Brennan if they are meaningful?
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=120mm, 
        scale=1]{Fig4.pdf}
        \caption{Flux diagram between PCCA+ states are shown for each sequences, accompanied by representative structures are for each macrostate. Arrows indicate the probability of transitioning out of the state within the lag time. Circle areas are proportional to state stability.}

@@

Following the analysis pipeline described above, we generated SRV-MSMs for each sequence. We identified seven kinetically relevant states that are captured within the resolution of the model. Figure \ref{fig:allseq_table} shows cartoon and molecular rendering for each of these states as well as their sequence-specific stationary probabilities. By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S1, 3S1, 5S2, 3S2) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D). Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models. Arrows between states show the conditional transition probability to another state within the MSM lag time. We now proceed to analyse the sequence-dependent thermodynamics and kinetic features of each MSM.

@@

Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends\citep{Santalucia2004TM}. The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models \citep{Michele2014EHybridization}. It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions. In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (figure \ref{fig:NN_table}). For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S1 and 5S2) are more energetically favorable than those in the 3' shifted state (3S1 and 3S2). We find qualitative agreement to our macrostate free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone. For GC-end, we consider C:T and G:A mismatches as non-interacting dangling ends such that each shifted conformation has four total dangling ends. Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.  Contrary to macrostate results, however, the 3S1 state is more stable than 5S1 states. Moreover, we do not observe a stable 3S2 cluster for GC-end. This indicates that 5' vs. 3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone \citep{Doktycz1990ThermodynamicATGC}.

@@

% Although NN calculations predict 5' vs. 3' stabilities for AT-all, we observe that GC-end 5S1 state remains more populated than the 3S1 state contrary to NN predictions. Indeed, we do not observe the formation of a stable 3S2 state for GC-end, further indicating that these states 3' states are unfavorable. This leads us to believe that 5' vs. 3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN effects \citep{Doktycz1990ThermodynamicATGC}.

@@

In addition to state probabilities and free energy approximations, the coarse-grained MSM yields valuable discrete kinetic information in the form of transition probabilities between states. For AT-all, we observe approximately equal probability of transitioning from D to any other state. Once a transition has been made, however, the 5' vs. 3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate. Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S1 $\rightarrow$ H is more favorable than 5S1 $\rightarrow$ 5S2 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S1 $\rightarrow$ D) than to more aligned states (5S1 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability 3S2 $\rightarrow$ D is 10x higher than the 3S2 $\rightarrow$ 3S1 path towards native hybridization.

For GC-end we observe no significant flux between the 5S2 state and the structurally similar 5S1 state, indicating that the former acts more as a kinetic trap than a pathway to H. The 5S1 and 3S1 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D when compared to AT-all. This suggests that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S1 $\rightarrow$ H and 3S1 $\rightarrow$ H once shifted states have formed. This phenomenon will be explored more in depth below.

@@

Taken together, our results show that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways. A majority of initial hybridization contacts occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways and 37\% of H $\rightarrow$ D pathways pass through some out-of-register state. The 3S2 and 5s2 states are more prone to re-dissociation and thus contribute to about 5\% of these pathways. While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through any out-of-register states. These observations show that small disruptions to repetitive tracts shift hybridization pathways towards more two-state behavior. Indeed, there remain some repetitive tracts in GC-mix and GC-core, however we observe no out-of-register pathways for these sequences. Furthermore, our results indicate that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}. These processes can occur even when terminal mismatches are present in the non-native structure.

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=150mm, scale=1]{Fig8.pdf}
        \caption{First two diffusion map coordinates built from 10000 5S1 states, equally sampled from AT-all and GC-end. Color maps show inverse distances (nm$^{-1}$) between out-of-register ends and complementary ends. The lower right region of the diffusion map space shows a high concentration of "shifted loop" base pairs that may facilitate the out-of-register GC-end transition.}
        \label{fig:GC-end_dmaps}
	\end{center}
\end{figure}

% need to gear this more towards the kinetics
\subsection{5s1 state distributions reveal differences between AT-all and GC-end kinetics}

Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates and during transition periods between states. These configurations interchange significantly faster than the SRV and MSM lag times, but we can use a diffusion map approach to glean a structural understanding of the macrostate population. Diffusion maps generate a low dimensional embedding of the data into high variance structural modes and are well-suited to find subtle differences in temporally disconnected data \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}.
% add something here on previous dmap refs?

While examining molecular renderings, we noticed that a significant proportion of GC-end 5s1 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register. We would not necessarily expect duplexes to sacrifice helical conformational entropy in order to facilitate termini bonding. To compare how these state populations differ between GC-end and AT-all, we employed diffusion maps built on an equal sampling of 5000 conformations from the 5S1 state of both sequences. We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation. This created degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating effects at the identical "top" and "bottom" of the strands. In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI. Diffusion maps built from samples of the 3S1 states are also shown in the SI.

The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair. As expected, this shows that mismatched C:T pairs are never bound whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region. The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, highlights structural variances within the two sequences distributions. We notice that about 40\% of GC-end configurations fall below $\Psi_3 = -0.01$  compared to only 10\% of AT-all configuration. We find that these GC-end populations share stabilizing termini contacts despite all A:T contacts being shifting out-of-register. These "shifted-loop" bonds appear uniquely stable for GC-end conformations,  and suggest an intermediate states through which hybridization is facilitated. This may account for the GC-end 5S1 $\rightarrow$ H transition being more favorable than  5S1 $\rightarrow$ D despite the state free energy being substantially higher than the AT-all equivalent.

@@

(I just completed most of this analysis on Friday. I think this shows a more direct comparison to Paul's 2016 kinetic model -- mainly all-or-nothing deviations caused by fraying are ordered GC-core > GC-mix > AT-all > GC-end. In addition we see a distinct state form for GC-core which serves to verify the F4 state and its kinetic relevance. Lastly it provides another means by which MSMs can inform analysis techniques for clearer comparison between sequences. I could repeat this procedure and analysis for direct dissociation as well if that is of interest)

To more directly compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$D for GC-core) events and excluded transitions that passed through long-lived out-of-register states. To achieve higher resolutions, we reran simulations with a 10x higher save rate and selected a 1000 frame region (10 ns simulation time) for each event. We built diffusion maps on a set of 48 such "trimmed" events evenly sampled across sequences. We found a distinct spectral gap after the first non-trivial diffusion map eigenvalue, indicating that the leading mode $\Psi_2$ contained most of the structural variance. Indeed, we found $\Psi_2$ had a 0.89 Pearson correlation with simulation time and a  -0.94 correlation with inter-strand core distance with minimal variation by sequence. This shows that $\Psi_2$ is strongly correlated with hybridization progress for all sequences and represents a consistent collective variable for comparison between sequences.

Once we determined a strong hybridization coordinate, we compared probability distributions for each sequence along $\Psi_2$ (Figure \ref{fig:Fig8}). The region where $\Psi_2 < 0$ is noisy and mainly reflects the distance between approaching strands, but we noted statistically meaningful differences between sequence populations once first WC contacts are made. In the $ 0.001 <\Psi_2 < 0.0035$ region, heatmaps show that contacts tend to form near the center of the duplex, and frayed structures progressively zipper closed as $\Psi_2$ increases. As expected, GC-core shows a distinct peak in the frayed region, corresponding to the F4 macrostate identified by SRV-MSMs. GC-mix has a notably higher population than AT-all and GC-end but displays a more gradual progression toward the native hybridized state. This is consistent with the downhill nucleation-zippering mechanism described above. Although differences between AT-all and GC-end are more subtle, overall sequence populations in this region follow predictions by Sanstead et. al. regarding deviations from two-state binding behavior, where GC-core > GC-mix > AT-all > GC-end \citep{Sanstead2016}. This qualitative agreement shows that MSM information can be adapted to compare sequences along common pathways and elucidate dynamics occurring faster than the lag time. Furthermore, we are able to distinguish out-of-register deviations from fraying deviations and describe relevant experimental comparisons for each.

@@

\subsection{Explicit model comparisons}

>>>>

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA}. Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides, Roman o2013DNADependence, Hinckley2014Coarse-grainedEffects}. Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}. Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.

@@

In this work, we study the same four sequences explored by Sanstead et al. in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}. We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}. We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent. We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.
% this section should look more like a conclusion foreshadowing conclusions and main take aways 
We find that 3spn2 allows us to access long length and time scales and observe sequence-dependent kinetics and relaxation rates that are, within a corrective scaling factor, in agreement with observed experimental trends. Applying SRV-MSMs enables the construction of high-resolution sequence-dependent kinetic models of hybridization and dissociation. Consistent with previous studies \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity}, we find the degree of repetitiveness in the sequence and therefore the kinetic accessibility and thermodynamic stability of shifted states leads to richer dynamics populated by a diversity of long-lived metastable states. Disrupting repetitive stretches of bases by judicious placement of interrupting G:C pairs can tune the landscape from rich 6-state to very simple 2-state (“all-or-nothing”) behavior. Furthermore, concentrating G:C pairs at the center of a duplex produces a longer-live frayed state that facilitates the hybridization and dissociation process. Analysis within and between metastable state populations using diffusion maps allows us to explore the conformational diversity and heterogeneity within each state at time scales shorter than the kinetic resolution of the SRV-MSMs. This provides us with molecular-level understanding for the differential stability of various conformational states as a function of oligomer sequence. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.

@@

We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm \citep{Sanstead2016}. We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff. We ran our simulations in the NVT ensemble. Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent. \citep{Schneider1978Molecular-dynamicsTransitions}. Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states. We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequence, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation. Half of runs were initialized from the native hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. We observed between 55-100 hybridization/dissociation events for each of the four sequences.

@@

\subsection{\label{sec:methods}Diffusion Maps}
Diffusion maps generate a low-dimensional non-linear embedding of data into high variance structural modes \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}. They are well-suited to find subtle differences in molecular conformations that are temporally disconnected or that interchange faster than the MSM lag time. We used reciprocal inter-molecular base distances as inputs to diffusion maps, and tuned $\epsilon$ and $\alpha$ parameters in accordance with previous studies \citep{Shmilovich, Wang2017MesoscaleLandscapes}. We found that the choice of these parameters mainly served to condense or stretch the low dimensional space and had a minimal qualitative impact on salient results.

@@

% change state names and scientific notation
% include D here? can see what it would look like...
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=160mm, 
        scale=1]{Fig3.pdf}
        \caption{Nearest neighbor representations, molecular renderings, and sequence-dependent probabilities for PCCA+ macrostates. Shifted state abbreviations indicate the direction of the shifted overhang (5' vs. 3') and the number of shifted base pairs relative the native state (2 vs. 4).}
        \label{fig:allseq_table}
	\end{center}
\end{figure}

% potentially get rid of State representation, ask Brennan if they are meaningful?
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=160mm, 
        scale=1]{Fig4.pdf}
        \caption{Flux diagram between PCCA+ states are shown for each sequences, accompanied by representative structures are for each macrostate. Arrows indicate the probability of transitioning out of the state within the lag time. Circle areas are proportional to state stability.}

@@

Following the analysis pipeline described above, we generated SRV-MSMs for each sequence. We identified seven kinetically relevant states that are captured within the resolution of the model. Figure \ref{fig:allseq_table} shows cartoon and molecular rendering for each of these states as well as their sequence-specific stationary probabilities. By row, these include the fully hybridized state (H) in which all native base pairings are intact, four "shifted" states (5S2, 3S2, 5S4, 3S4) where complementary base pairs form out-of-register, a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound, and the fully dissociated state (D). Figure \ref{fig:allseq_transition} shows how these states fit into sequence-specific low-dimensional representations and kinetic models. Arrows between states show the conditional transition probability to another state within the MSM lag time. We now proceed to analyse the sequence-dependent thermodynamics and kinetic features of each MSM.

@@

Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends\citep{Santalucia2004TM}. The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models \citep{Michele2014EHybridization}. It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions. In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (figure \ref{fig:NN_table}). For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S2 and 5S4) are more energetically favorable than those in the 3' shifted state (3S2 and 3S4). We find qualitative agreement to our macrostate free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone. For GC-end, we consider C:T and G:A mismatches as non-interacting dangling ends such that each shifted conformation has four total dangling ends. Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.  Contrary to macrostate results, however, the 3S2 state is more stable than 5S2 states. Moreover, we do not observe a stable 3S4 cluster for GC-end. This indicates that 5' vs. 3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone \citep{Doktycz1990ThermodynamicATGC}.

@@

% Although NN calculations predict 5' vs. 3' stabilities for AT-all, we observe that GC-end 5S2 state remains more populated than the 3S2 state contrary to NN predictions. Indeed, we do not observe the formation of a stable 3S4 state for GC-end, further indicating that these states 3' states are unfavorable. This leads us to believe that 5' vs. 3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN effects \citep{Doktycz1990ThermodynamicATGC}.

@@

In addition to state probabilities and free energy approximations, the coarse-grained MSM yields valuable discrete kinetic information in the form of transition probabilities between states. For AT-all, we observe approximately equal probability of transitioning from D to any other state. Once a transition has been made, however, the 5' vs. 3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate. Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S2 $\rightarrow$ H is more favorable than 5S2 $\rightarrow$ 5S4 -- suggesting that more shifted states play a greater role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S2 $\rightarrow$ D) than to more aligned states (5S2 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability 3S4 $\rightarrow$ D is 10x higher than the 3S4 $\rightarrow$ 3S2 path towards native hybridization.

For GC-end we observe no significant flux between the 5S4 state and the structurally similar 5S2 state, indicating that the former acts more as a kinetic trap than a pathway to H. The 5S2 and 3S2 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D when compared to AT-all. This suggests that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S2 $\rightarrow$ H and 3S2 $\rightarrow$ H once shifted states have formed. This phenomenon will be explored more in depth below.

@@

Taken together, our results show that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways. A majority of initial hybridization contacts occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways and 37\% of H $\rightarrow$ D pathways pass through some out-of-register state. The 3S4 and 5S4 states are more prone to re-dissociation and thus contribute to about 5\% of these pathways. While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through any out-of-register states. These observations show that small disruptions to repetitive tracts shift hybridization pathways towards more two-state behavior. Indeed, there remain some repetitive tracts in GC-mix and GC-core, however we observe no out-of-register pathways for these sequences. Furthermore, our results indicate that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}. These processes can occur even when terminal mismatches are present in the non-native structure.

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=150mm, scale=1]{Fig8.pdf}
        \caption{First two diffusion map coordinates built from 10000 5S2 states, equally sampled from AT-all and GC-end. Color maps show inverse distances (nm$^{-1}$) between out-of-register ends and complementary ends. The lower right region of the diffusion map space shows a high concentration of "shifted loop" base pairs that may facilitate the out-of-register GC-end transition.}
        \label{fig:GC-end_dmaps}
	\end{center}
\end{figure}

% need to gear this more towards the kinetics
\subsection{5S2 state distributions reveal differences between AT-all and GC-end kinetics}

Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates and during transition periods between states. These configurations interchange significantly faster than the SRV and MSM lag times, but we can use a diffusion map approach to glean a structural understanding of the macrostate population \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}. While examining molecular renderings, we noticed that a significant proportion of GC-end 5S2 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register. We would not necessarily expect duplexes to sacrifice helical conformational entropy in order to facilitate termini bonding. To compare how these state populations differ between GC-end and AT-all, we employed diffusion maps built on an equal sampling of 5000 conformations from the 5S2 state of both sequences. We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation. This created degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating effects at the identical "top" and "bottom" of the strands. In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI. Diffusion maps built from samples of the 3S2 states are also shown in the SI.

The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair. As expected, this shows that mismatched C:T pairs are never bound whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region. The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, highlights structural variances within the two sequences distributions. We notice that about 40\% of GC-end configurations fall below $\Psi_3 = -0.01$  compared to only 10\% of AT-all configuration. We find that these GC-end populations share stabilizing termini contacts despite all A:T contacts being shifting out-of-register. These "shifted-loop" bonds appear uniquely stable for GC-end conformations,  and suggest an intermediate states through which hybridization is facilitated. This may account for the GC-end 5S2 $\rightarrow$ H transition being more favorable than  5S2 $\rightarrow$ D despite the state free energy being substantially higher than the AT-all equivalent.

@@

To compare fast hybridization dynamics across all sequences, we trimmed simulations centered on direct D$\rightarrow$H (or D$\rightarrow$F4$\rightarrow$D for GC-core) events and excluded transitions that passed through long-lived out-of-register states. To achieve higher resolutions, we reran simulations with a 10x higher save rate and selected a 1000 frame region (10 ns simulation time) for each event. We built diffusion maps on a set of 48 such "trimmed" events evenly sampled across sequences. We found a distinct spectral gap after the first non-trivial diffusion map eigenvalue, indicating that the leading mode $\Psi_2$ contained most of the structural variance. Indeed, we found $\Psi_2$ had a 0.89 Pearson correlation with simulation time and a  -0.94 correlation with inter-strand core distance with minimal variation by sequence. This shows that $\Psi_2$ is strongly correlated with hybridization progress for all sequences and represents a consistent collective variable for comparison between sequences.

Once we determined a promising hybridization coordinate, we compared probability distributions for each sequence along $\Psi_2$ (Figure \ref{fig:Fig8}). The region where $\Psi_2 < 0$ is noisy and mainly reflects the distance between approaching strands, but we noted statistically meaningful differences between sequence populations once first WC contacts are made. In the $ 0.001 <\Psi_2 < 0.0035$ region, heatmaps show that contacts tend to form near the center of the duplex, and frayed structures progressively zipper closed as $\Psi_2$ increases. As expected, GC-core shows a distinct peak in the frayed region, corresponding to the F4 macrostate identified by SRV-MSMs. GC-mix has a notably higher population than AT-all and GC-end but displays a more gradual progression toward the native hybridized state. This is consistent with the downhill nucleation-zippering mechanism described above. Although differences between AT-all and GC-end are more subtle, overall sequence populations in this region follow predictions by Sanstead et. al. regarding deviations from two-state binding behavior, where GC-core > GC-mix > AT-all > GC-end \citep{Sanstead2016}. This qualitative agreement shows that MSM information can be adapted to compare sequences along common pathways and elucidate dynamics occurring faster than the lag time. Furthermore, we are able to distinguish out-of-register deviations from fraying deviations and describe relevant experimental comparisons for each.

@@

%\subsection{Explicit model comparisons}
